# Motivation for confidence interval

Let us consider a Gaussian distribution with mean 0 and variance 2:  
$$
  X \sim \mathcal{N}(0,2)
$$

[**Simulating the sample mean of a Gaussian distribution multiple times **]{.lead}  
[See the rnorm section](The_normal_distribution.qmd#sec-rnorm-function)
```{r}

M <- c() # initialize an empty vector to store the means

# Loop over three different sample sizes: 50, 500, and 5,000 
for (k in c(50,500,5000)) 
{
    for (i in 1:50)             # simulate the sample mean 50 times    
    {
        A <- rnorm(k,0,sqrt(2)) # generate k observations with mean = 0, variance = 2
        m <- mean(A)
        M <- c(M,m)
    }
}
```

We now create a data frame D with:  
- a column **mean** containing the 150 sample means (50 for each sample size), and  
- a column **label** indicating the corresponding sample size.  

```{r}
D <- data.frame(mean = M, label = rep(c('50','500','5000'), each = 50))  # 'rep(..., each = 50)' aligns labels with sample sizes
boxplot(D$mean ~ D$label,
        main = "Distribution of Sample Means by Sample Size",
        xlab = "Sample Size", ylab = "Sample Mean")
```

::: {.callout-note title="Professor's Insight: Accuracy and Sample Size"}
As we can see in the boxplot:  
- For a sample size of 50, the sample means fluctuate roughly between -0.4 and 0.4.
- With 500 observations, the range narrows to around -0.2 to 0.1.
- And for 5,000 observations, the fluctuation becomes minimal, between -0.08 and 0.05.

**This illustrates that as the sample size increases, the accuracy of the sample mean improves**, because the fluctuations (i.e., variance) decrease.
The quality of the estimation improves at a rate proportional to 1/√n.
:::

