# Simple Linear Regression

with simulated dataset :

```{r}
# Creation of variables for two variables 
# X as 100 observation from -7 to 10 that follow uniform law 
X <- runif(100,-7,10)
# Y linearly 'connected' to X + some gaussian noise
Y <- - 4 + 2 * X + rnorm(100,0,2)
# Plot of the cloud of point
plot(X,Y)
```


```{r}

L <- lm(Y~X) # perform a linea model 
plot(X,Y)
# create a red line that fit the linear regression
abline(L,col='red') # equal abline(reg= L, col='red'
```

 The red line in the plot above represents the fitted simple linear regression model (Y ~ X).  
 It shows **the best linear relationship between X and Y** based on the simulated data.

 we create a linear link as  
 $$
 Y= -4 + 2 \times X + \mathcal{N}(0;2)
 $$
 
 Thus we expect a coefficient director of 2 and an intercept at 4 : 


```{r}
print(L)
```

> The equation of L is $-3.686 + 1.962X$ because it is estimation thanks to observations, we are not going to be able to exactly get the theoritical model, but an estimation.    