# The Tidyverse

```{r}
library(tidyverse)
library(dslabs)
data(murders)
```

# Manipulating Data frames with dplyr and purrr

## Tidy Data

We say that a data table is in *tidy* format if each row represents one observation and columns
represent the different variables available for each of these observations.
The murders dataset is an example of a tidy data frame.

```{r}
head(murders)
```

Each row represents a state with each of the five columns providing a different variable related
to these states: name, abbreviation, region, population, and total murders.

## Manipulating Data Frames

>"The dplyr package from the tidyverse introduces functions that perform some of the most
common operations when working with data frames and uses names for these functions that
are relatively easy to remember. For instance, to change the data table by adding a new
column, we use mutate. To filter the data table to a subset of rows, we use filter. Finally,
to subset the data by selecting specific columns, we use select."


### The `mutate` function
The `mutate` function is used to add new columns to a data frame or modify existing ones.

```{r}
# Add a new column 'rate' to the murders data frame
murders  <- mutate(murders, rate = total / population * 100000)
```

**Note:** to compute the rate, we used `total` and `population` columns, which are not defined in the
global environment. The `mutate` function allows us to use the names of the columns directly.

>"This is one of dplyr's main features. Functions in this package, such as mutate, know to look for variables in th data frame provided in the first argument.  
In the call to mutate above, `total` will have the values in `murders$total`. This approach makes the code much more readable and concise."


```{r}
head(murders)
```

**Note**: the `mutate` function does not change the original data frame.  

>"Although we have overwritten the original **murders** object, this does not change the object that is loaded with `data(murders)`.  
If we load the murders data again, the original will overwritte our mutated version."

### Subsetting with `filter`
The `filter` function is used to subset rows based on logical conditions.

*Filter the murders data frame to include only the entries for which the murder rate is lower than 0.71.*

```{r}
# Syntax : data, conditional statement.
filter(murders, rate <= 0.71)
```

### Selecting columns with `select`

The `select()` function is used to extract specific columns from a data frame.

In the example below:
- We create a new data frame containing only the columns state, region, and rate.
- We then apply `filter()` to keep only the rows where the murder rate is less than or equal to 0.71.

```{r}
state_region_rate_table <- select(murders, state, region, rate)
filter(state_region_rate_table, rate <= 0.71)
```

### Exercises

1. Load the dplyr package and the murders dataset.
```{r}
library(dplyr)
library(dslabs)
data(murders)
```

2. Use the function mutate to add a column rank containing the rank,from highest to lowest murder rate.  
Make sure you redefine murders so we can keep using this variable.
```{r}
murders <- mutate(murders, rate = total / population * 10^5)
murders <- mutate(murders, rank = rank(-rate))
murders %>% head()
select(murders, state, population) %>% head()
```
*We can write population rather than murders$population. The function mutate knows we are grabbing columns from murders.*


3. Use `Select` to show the state names and abbreviations in murders. Do not redefine murders, just show the results.

```{r}
select(murders, state, abb)
```

4. Use filter to show the top 5 states with the highest murder rates.

```{r}
filter(murders,  rank <= 5)
```

5. Create a new data frame called no_south that removes states from the South region. How
many states are in this category? You can use the function nrow for this.

**Note**: We can remove rows using the != operator. For example, to remove Florida, we would do
this:
```{r}
no_florida <- filter(murders, state != "Florida")
```

```{r}
# Create the new data frame without south region
no_south <- filter(murders, region != "South")
# Compute how many states are not in the south
select(no_south, state) %>% nrow()
```

*There are 34 states which are not in the south*

